---
title: "figure-RG"
author: "Harry Robertson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 14,
  fig.height = 7,
  dpi = 600
)

library(tidyverse)
library(patchwork)
library(scales)
library(ggpubr)

RESULTS <- "results"

# Main-text figure: composite radar panel (this Rmd's headline output).
FIG_DIR <- "main-figures"
# Supplementary figures: per-metric bar+CI panels go to dataset-specific dirs.
SUPP_MIMIC_DIR <- file.path("supp-figures", "MIMIC-CXR-figures")
SUPP_IU_DIR    <- file.path("supp-figures", "IU-xray-figures")

walk(c(FIG_DIR, SUPP_MIMIC_DIR, SUPP_IU_DIR),
     ~ dir.create(.x, showWarnings = FALSE, recursive = TRUE))

N_BOOT <- 10000
SEED   <- 42

source("./figure2-helpers.R")
```

```{r data}
RG_NLG_METRICS <- c("bleu1", "bleu2", "bleu4", "meteor", "rougel", "cider")
RG_SEM_METRICS <- c("bertscore", "semb_score", "radgraph_combined")
RG_CE_METRICS  <- c("CE_precision",  "CE_recall",  "CE_f1",
                    "CE5_precision", "CE5_recall", "CE5_f1")
RG_RADAR_METRICS <- c(RG_NLG_METRICS, RG_SEM_METRICS, RG_CE_METRICS)

rg_mimic_ps      <- load_per_sample(file.path(RESULTS, "performance/rg/mimic_cxr"))
rg_mimic_summary <- build_summary(rg_mimic_ps, RG_RADAR_METRICS)
rg_mimic_boot    <- compute_boot_cis(rg_mimic_ps, RG_RADAR_METRICS)

rg_iu_ps      <- load_per_sample(file.path(RESULTS, "performance/rg/iu_xray"))
rg_iu_summary <- build_summary(rg_iu_ps, RG_RADAR_METRICS)
rg_iu_boot    <- compute_boot_cis(rg_iu_ps, RG_RADAR_METRICS)
```

# Main figure — RG radars (MIMIC-CXR + IU X-ray)

```{r figure-RG-main, fig.width=14, fig.height=7}
p_mimic <- radar_plot(
  rg_mimic_summary,
  RG_RADAR_METRICS,
  title = "MIMIC-CXR",
  axis_limits = MIMIC_CXR_RADAR_LIMITS,
  label_size = 2.6,
  grid_label_size = 2.1
)
p_iu    <- radar_plot(rg_iu_summary,    RG_RADAR_METRICS, title = "IU X-ray")

figure_rg <- (p_mimic + p_iu) +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "bottom",
    plot.tag        = element_text(face = "bold", size = 14)
  )

ggsave(file.path(FIG_DIR, "figure-RG.pdf"),
       plot = figure_rg, width = 14, height = 7, device = "pdf")

figure_rg
```

# Supplementary panels

Each per-metric bar + 95% bootstrap CI is saved individually to
`supp-figures/{MIMIC-CXR,IU-xray}-figures/`, plus a 3-row arranged composite
per metric family.

```{r supp-helpers}
save_supp_panels <- function(boot_df, metric_specs, supp_dir,
                             y_limits = c(0, 1)) {
  imap(metric_specs, function(metric, figure_name) {
    p <- bar_with_ci(boot_df, metric) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    if (!is.null(y_limits)) {
      p <- p + scale_y_continuous(limits = y_limits,
                                  breaks = seq(y_limits[1], y_limits[2], 0.2))
    }
    ggsave(file.path(supp_dir, paste0(figure_name, ".pdf")),
           plot = p, width = 5, height = 4, device = "pdf")
    p
  })
}

save_supp_arranged <- function(plots, filename, supp_dir,
                               ncol, nrow, width, height,
                               legend = "bottom") {
  arr <- ggarrange(plotlist = plots, ncol = ncol, nrow = nrow,
                   common.legend = TRUE, legend = legend)
  ggsave(file.path(supp_dir, filename),
         plot = arr, width = width, height = height, device = "pdf")
  arr
}
```

## MIMIC-CXR — NLG metrics (a–f)

```{r supp-mimic-nlg}
mimic_nlg_specs <- c(
  figure2a_bleu1  = "bleu1",
  figure2b_bleu2  = "bleu2",
  figure2c_bleu4  = "bleu4",
  figure2d_meteor = "meteor",
  figure2e_rougel = "rougel",
  figure2f_cider  = "cider"
)
mimic_nlg_plots <- save_supp_panels(rg_mimic_boot, mimic_nlg_specs,
                                    SUPP_MIMIC_DIR, y_limits = NULL)
save_supp_arranged(mimic_nlg_plots, "figure2a-f.pdf", SUPP_MIMIC_DIR,
                   ncol = 3, nrow = 2, width = 9, height = 7)
```

## MIMIC-CXR — Semantic metrics (g–i)

```{r supp-mimic-sem}
mimic_sem_specs <- c(
  figure2g_bert     = "bertscore",
  figure2h_semb     = "semb_score",
  figure2i_radgraph = "radgraph_combined"
)
mimic_sem_plots <- save_supp_panels(rg_mimic_boot, mimic_sem_specs, SUPP_MIMIC_DIR)
save_supp_arranged(mimic_sem_plots, "figure2g-h.pdf", SUPP_MIMIC_DIR,
                   ncol = 3, nrow = 1, width = 9, height = 3.5, legend = "none")
```

## MIMIC-CXR — Clinical efficacy (j–o)

```{r supp-mimic-ce}
mimic_ce_specs <- c(
  figure2j_ce_precision  = "CE_precision",
  figure2k_ce_recall     = "CE_recall",
  figure2l_ce_f1         = "CE_f1",
  figure2m_ce5_precision = "CE5_precision",
  figure2n_ce5_recall    = "CE5_recall",
  figure2o_ce5_f1        = "CE5_f1"
)
mimic_ce_plots <- save_supp_panels(rg_mimic_boot, mimic_ce_specs, SUPP_MIMIC_DIR)
save_supp_arranged(mimic_ce_plots, "figure2j-o.pdf", SUPP_MIMIC_DIR,
                   ncol = 3, nrow = 2, width = 9, height = 7)
```

## MIMIC-CXR — Selected 3x3 composite (a-h, o)

```{r supp-mimic-selected-3x3}
mimic_selected_plots <- c(
  mimic_nlg_plots,
  mimic_sem_plots[c("figure2g_bert", "figure2h_semb")],
  mimic_ce_plots["figure2o_ce5_f1"]
)

save_supp_arranged(mimic_selected_plots, "figure2a-h_o.pdf", SUPP_MIMIC_DIR,
                   ncol = 3, nrow = 3, width = 9, height = 9)
```

## IU X-ray — NLG metrics (a–f)

```{r supp-iu-nlg}
iu_nlg_specs <- c(
  figure3a_bleu1  = "bleu1",
  figure3b_bleu2  = "bleu2",
  figure3c_bleu4  = "bleu4",
  figure3d_meteor = "meteor",
  figure3e_rougel = "rougel",
  figure3f_cider  = "cider"
)
iu_nlg_plots <- save_supp_panels(rg_iu_boot, iu_nlg_specs,
                                 SUPP_IU_DIR, y_limits = NULL)
save_supp_arranged(iu_nlg_plots, "figure3a-f.pdf", SUPP_IU_DIR,
                   ncol = 3, nrow = 2, width = 9, height = 7)
```

## IU X-ray — Semantic metrics (g–i)

```{r supp-iu-sem}
iu_sem_specs <- c(
  figure3g_bert     = "bertscore",
  figure3h_semb     = "semb_score",
  figure3i_radgraph = "radgraph_combined"
)
iu_sem_plots <- save_supp_panels(rg_iu_boot, iu_sem_specs, SUPP_IU_DIR)
save_supp_arranged(iu_sem_plots, "figure3g-i.pdf", SUPP_IU_DIR,
                   ncol = 3, nrow = 1, width = 9, height = 3.5, legend = "none")
```

## IU X-ray — Clinical efficacy (j–o)

```{r supp-iu-ce}
iu_ce_specs <- c(
  figure3j_ce_precision  = "CE_precision",
  figure3k_ce_recall     = "CE_recall",
  figure3l_ce_f1         = "CE_f1",
  figure3m_ce5_precision = "CE5_precision",
  figure3n_ce5_recall    = "CE5_recall",
  figure3o_ce5_f1        = "CE5_f1"
)
iu_ce_plots <- save_supp_panels(rg_iu_boot, iu_ce_specs, SUPP_IU_DIR)
save_supp_arranged(iu_ce_plots, "figure3j-o.pdf", SUPP_IU_DIR,
                   ncol = 3, nrow = 2, width = 9, height = 7)
```
