---
title: "figure-SLAKE-ablation-radar"
author: "Harry Robertson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 4.13,
  fig.height = 5.12,
  dpi = 600
)

library(tidyverse)
library(patchwork)
library(scales)

RESULTS  <- "results"

# Bootstrap parameters used by boot_mean_ci() in figure2-helpers.R.
N_BOOT <- 10000
SEED <- 42

source("./figure2-helpers.R")

MODEL_PALETTE <- c(
  MODEL_PALETTE,
  "Base"           = unname(MODEL_PALETTE[["Qwen2.5-VL 7B"]]),
  "Base + CoT"     = unname(MODEL_PALETTE[["Qwen3-VL 8B"]]),
  "Base + VReason" = unname(MODEL_PALETTE[["VReason"]])
)

ablation_label <- function(stem) {
  case_when(
    stem == "qwen2.5_vl_7b_instruct"     ~ "Base",
    stem == "qwen2.5_vl_7b_instruct_cot" ~ "Base + CoT",
    stem %in% c("vreason", "vreasson")   ~ "Base + VReason",
    TRUE ~ model_label(stem)
  )
}

load_ablation_per_sample <- function(directory) {
  csvs <- list.files(directory, pattern = "\\.csv$", full.names = TRUE)
  csvs <- csvs[!grepl("summary\\.csv$", csvs)]

  frames <- map(csvs, function(p) {
    stem <- tools::file_path_sans_ext(basename(p))
    df <- read_csv(p, show_col_types = FALSE)
    df$model <- stem
    df$model_label <- ablation_label(stem)
    df
  })
  bind_rows(frames)
}
```

```{r helpers}
VQA_METRIC_LABELS <- c(
  token_f1       = "Token F1",
  exact_match    = "Exact match",
  contains_match = "Contains match",
  token_recall   = "Token recall"
)

MODALITY_LEVELS <- c("CT", "MRI", "X-Ray")
ORGAN_LEVELS <- c(
  "Abdomen", "Brain", "Brain_Face", "Brain_Tissue",
  "Chest_mediastinal", "Lung", "Neck", "Pelvic Cavity"
)
CONTENT_TYPE_LEVELS <- c(
  "Abnormality", "Color", "KG", "Modality", "Organ",
  "Plane", "Position", "Quantity", "Shape", "Size"
)

pretty_stratum <- function(x) {
  str_replace_all(x, "_", " ")
}

# Stratification specs used by make_combined_metric_summary. Keeping these
# isolated lets us mix-and-match axes (modality + location, or
# content_type + modality + location for subcategory radars) without
# duplicating the bind_rows boilerplate.
STRATIFICATION_SPECS <- list(
  modality     = list(col = "modality",     levels = MODALITY_LEVELS,
                      transform = identity),
  location     = list(col = "location",     levels = pretty_stratum(ORGAN_LEVELS),
                      transform = pretty_stratum),
  content_type = list(col = "content_type", levels = CONTENT_TYPE_LEVELS,
                      transform = identity)
)

safe_filename <- function(x) {
  x |>
    str_replace_all("[^A-Za-z0-9]+", "_") |>
    str_replace_all("^_|_$", "") |>
    str_to_lower()
}

metric_label <- function(metric) {
  label <- VQA_METRIC_LABELS[[metric]]
  if (is.null(label)) str_replace_all(metric, "_", " ") else label
}

metric_filename <- function(metric) {
  metric |>
    str_replace_all("_", "-") |>
    str_replace_all("[^A-Za-z0-9-]+", "-") |>
    str_replace_all("^-|-$", "") |>
    str_to_lower()
}

make_combined_metric_summary <- function(ps_df, metric,
                                          strata = c("modality", "location"),
                                          comparison_labels = NULL) {
  specs <- STRATIFICATION_SPECS[strata]
  comparison_levels <- unlist(map(specs, "levels"), use.names = FALSE)

  if (is.null(comparison_labels)) {
    comparison_labels <- comparison_levels
  }
  label_lookup <- set_names(comparison_labels, comparison_levels)

  comparison_df <- map_dfr(specs, function(s) {
    tibble(
      model_label = ps_df$model_label,
      comparison  = s$transform(ps_df[[s$col]]),
      value       = ps_df[[metric]]
    )
  })

  summary_df <- comparison_df |>
    mutate(comparison = factor(comparison, levels = comparison_levels)) |>
    filter(!is.na(comparison)) |>
    group_by(model_label, comparison) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") |>
    mutate(
      model_label = factor(
        model_label,
        levels = c("Base", "Base + CoT", "Base + VReason")
      ),
      comparison = as.character(comparison),
      comparison_label = unname(label_lookup[comparison])
    ) |>
    arrange(model_label) |>
    select(-comparison) |>
    pivot_wider(names_from = comparison_label, values_from = value) |>
    mutate(model_label = as.character(model_label))

  # Drop comparison labels with no data after filtering (e.g. Shape/Quantity
  # have zero CLOSED questions) so the radar doesn't try to render NA axes.
  present_labels <- comparison_labels[comparison_labels %in% names(summary_df)]

  list(summary = summary_df, levels = present_labels)
}

# Per-spoke axis ranges for the SLAKE radars, matching the RG radars
# (MIMIC/IU), where each spoke is scaled to its own [min, max] instead of a
# shared 0-1 scale. The RG radars use a hand-tuned limits list; the SLAKE
# axes are dynamic categories (modality / organ / content-type), so we derive
# the limits from the data. For each axis we take the min/max across the three
# ablation variants, pad outward, then round to clean 0.05 bounds. Rounding
# outward guarantees every observed value sits strictly inside its range, so
# radar_plot()'s out-of-range guard never trips.
compute_radar_axis_limits <- function(summary_df, labels,
                                      pad_frac = 0.1, min_pad = 0.02,
                                      accuracy = 0.05) {
  set_names(
    lapply(labels, function(lbl) {
      vals <- summary_df[[lbl]]
      vals <- vals[is.finite(vals)]
      if (length(vals) == 0) return(c(0, 1))

      lo <- min(vals)
      hi <- max(vals)
      pad <- max(min_pad, pad_frac * (hi - lo))

      lo_p <- max(0, floor((lo - pad) / accuracy) * accuracy)
      hi_p <- min(1, ceiling((hi + pad) / accuracy) * accuracy)

      if (hi_p <= lo_p) {
        lo_p <- max(0, lo_p - accuracy)
        hi_p <- min(1, max(hi_p + accuracy, lo_p + accuracy))
      }
      c(lo_p, hi_p)
    }),
    labels
  )
}

make_ablation_metric_plot <- function(ps_df, metric, title,
                                      strata = c("modality", "location"),
                                      comparison_labels = NULL,
                                      label_size = 3.2, grid_label_size = 2.8,
                                      line_width = 1.1, point_size = 1.8,
                                      coord_xlim = c(-1.36, 1.36),
                                      coord_ylim = c(-1.16, 1.16),
                                      plot_margin = margin(4, 8, 4, 8),
                                      scale_axes = TRUE) {
  combined_metric <- make_combined_metric_summary(ps_df, metric, strata,
                                                  comparison_labels)

  axis_limits <- if (scale_axes) {
    compute_radar_axis_limits(combined_metric$summary, combined_metric$levels)
  } else {
    NULL
  }

  radar_plot(
    combined_metric$summary,
    combined_metric$levels,
    title = title,
    axis_limits = axis_limits,
    label_size = label_size,
    grid_label_size = grid_label_size,
    fill_alpha = 0.2,
    line_width = line_width,
    point_size = point_size
  ) +
    coord_equal(xlim = coord_xlim, ylim = coord_ylim, expand = FALSE, clip = "off") +
    guides(colour = guide_legend(nrow = 1, byrow = TRUE)) +
    theme(
      plot.title = element_text(size = 9, face = "bold", hjust = 0.5),
      legend.text = element_text(size = 6),
      legend.key.width = unit(0.32, "cm"),
      legend.key.height = unit(0.24, "cm"),
      legend.spacing.x = unit(0.08, "cm"),
      legend.margin = margin(0, 0, 0, 0),
      plot.margin = plot_margin
    )
}

make_ablation_token_f1_plot <- function(ps_df, title) {
  make_ablation_metric_plot(ps_df, "token_f1", title)
}

```

```{r data}
abl_vqa_slake_ps <- load_ablation_per_sample(file.path(RESULTS, "ablation/vqa/slake"))
colnames(abl_vqa_slake_ps)
abl_slake_summary <- build_summary(abl_vqa_slake_ps, VQA_METRIC_LABELS)
p_abl_mimic <- radar_plot(abl_slake_summary, VQA_METRIC_LABELS, title = "SLAKE (VQA ablation)")
```

# SLAKE Ablation Metric Radars

These plots compare Base, Base + CoT, and Base + VReason on SLAKE VQA metrics
across all modality and organ/body-region comparisons.

```{r slake-ablation-metric-radars, fig.width=4.13, fig.height=5.12}
ablation_metrics <- c("token_f1", "exact_match", "contains_match", "token_recall")

slake_ablation_metric_plots <- ablation_metrics |>
  set_names() |>
  map(\(metric) {
    make_ablation_metric_plot(
      abl_vqa_slake_ps,
      metric,
      paste0("SLAKE ablation ", metric_label(metric), " - all questions")
    )
  })

slake_ablation_token_f1_all <- slake_ablation_metric_plots[["token_f1"]]

walk(slake_ablation_metric_plots, print)
```

# SLAKE Ablation Token F1 / Token Recall Radars by Answer Type

These use the same axes and styling, split by `answer_type`.

```{r slake-ablation-token-metrics-by-answer}
answer_type_metric_plots <- crossing(
  metric = c("token_f1", "token_recall"),
  answer_type_level = c("OPEN", "CLOSED")
) |>
  pmap(function(metric, answer_type_level) {
    sub_ps <- abl_vqa_slake_ps |> filter(answer_type == answer_type_level)

    make_ablation_metric_plot(
      sub_ps,
      metric,
      paste0("SLAKE ablation ", metric_label(metric),
             " - ", answer_type_level, " questions")
    )
  })

walk(answer_type_metric_plots, print)
```

# SLAKE Ablation Subcategory Radars (content_type + modality + location)

One radar per (metric × answer-type slice) combining all three SLAKE
stratifications onto a single axis. Axes whose categories have no questions
in a given slice (e.g. `Shape`, `Quantity` for `CLOSED`) are dropped
automatically.

```{r slake-ablation-subcategory-radars, fig.width=7, fig.height=7}
subcategory_strata <- c("content_type", "modality", "location")

answer_type_slices <- list(
  all    = list(label = "all questions",    filter = function(df) df),
  open   = list(label = "OPEN questions",   filter = function(df) filter(df, answer_type == "OPEN")),
  closed = list(label = "CLOSED questions", filter = function(df) filter(df, answer_type == "CLOSED"))
)

subcategory_plots <- crossing(
  metric = ablation_metrics,
  slice  = names(answer_type_slices)
) |>
  pmap(function(metric, slice) {
    slice_spec <- answer_type_slices[[slice]]
    sub_ps <- slice_spec$filter(abl_vqa_slake_ps)

    make_ablation_metric_plot(
      sub_ps,
      metric,
      paste0("SLAKE subcategory ", metric_label(metric),
             " - ", slice_spec$label),
      strata          = subcategory_strata,
      label_size      = 2.6,
      grid_label_size = 2.4,
      line_width      = 0.9,
      point_size      = 1.4,
      coord_xlim      = c(-1.55, 1.55),
      coord_ylim      = c(-1.4, 1.4),
      plot_margin     = margin(6, 18, 6, 18)
    )
  })

walk(subcategory_plots, print)
```

# SLAKE Ablation Content-Type + Region Radars per Modality

One radar per imaging modality (CT, MRI, X-Ray). Each radar subsets to all
SLAKE images of that modality and uses both the `content_type` categories and
the organ/body-region (`location`) strata as axes, comparing Base / Base + CoT
/ Base + VReason on `token_recall`. This drills the combined modality+organ
radar down so that each modality gets its own radar over the remaining findings
(content types and regions). Axes with no questions in a given modality (e.g.
`Shape` outside CT, or regions that only appear under one modality) are dropped
automatically.

```{r slake-ablation-modality-content-type-region-radars, fig.width=13.5, fig.height=5.5}
modality_radar_metric <- "token_recall"

make_modality_findings_radar <- function(ps_df, modality_level, metric,
                                          title = NULL) {
  sub_ps <- ps_df |> filter(modality == modality_level)
  if (is.null(title)) {
    n_examples <- n_distinct(sub_ps$id)
    title <- paste0(modality_level, " (", metric_label(metric),
                    ", n = ", n_examples, ")")
  }

  make_ablation_metric_plot(
    sub_ps, metric, title,
    strata          = c("content_type", "location"),
    label_size      = 2.6,
    grid_label_size = 2.4,
    line_width      = 0.9,
    point_size      = 1.4,
    coord_xlim      = c(-1.55, 1.55),
    coord_ylim      = c(-1.4, 1.4),
    plot_margin     = margin(6, 18, 6, 18)
  )
}

modality_findings_radars <- MODALITY_LEVELS |>
  set_names() |>
  map(\(mod) {
    make_modality_findings_radar(abl_vqa_slake_ps, mod,
                                 modality_radar_metric)
  })

modality_findings_panel <- wrap_plots(modality_findings_radars, ncol = 3,
                                      guides = "collect") +
  plot_annotation(
    title = paste0("SLAKE ablation by content type + region, per modality - ",
                   metric_label(modality_radar_metric)),
    theme = theme(plot.title = element_text(size = 11, face = "bold",
                                            hjust = 0.5))
  ) &
  theme(legend.position = "bottom")

modality_findings_panel
```

# SLAKE Ablation Bar Plots with Bootstrap CIs (split by stratum)

One bar plot per (metric × stratification × answer-type slice) comparing
Base / Base + CoT / Base + VReason. y-axis is the metric, x-axis is the
stratum level (e.g. CT / MRI / X-Ray for modality), bars are dodged by
ablation variant with 95% bootstrap CI error bars (B = 10,000).

```{r slake-ablation-bar-helpers}
compute_strata_boot_cis <- function(ps_df, metrics, stratum_col,
                                    stratum_levels = NULL,
                                    pretty = FALSE) {
  if (is.null(stratum_levels)) {
    stratum_levels <- sort(unique(ps_df[[stratum_col]]))
  }
  display_levels <- if (pretty) pretty_stratum(stratum_levels) else stratum_levels

  ps_df |>
    select(model_label, all_of(stratum_col), all_of(metrics)) |>
    rename(stratum = !!stratum_col) |>
    mutate(stratum = factor(stratum, levels = stratum_levels)) |>
    filter(!is.na(stratum)) |>
    pivot_longer(all_of(metrics), names_to = "metric", values_to = "value") |>
    group_by(model_label, stratum, metric) |>
    summarise(boot_mean_ci(value), .groups = "drop") |>
    mutate(
      stratum = factor(
        if (pretty) pretty_stratum(as.character(stratum)) else as.character(stratum),
        levels = display_levels
      ),
      model_label = factor(
        model_label,
        levels = c("Base", "Base + CoT", "Base + VReason")
      )
    )
}

bar_with_ci_by_stratum <- function(boot_df, metric_name, title,
                                   y_label = NULL,
                                   y_limits = c(0, 1)) {
  plot_df <- boot_df |> filter(metric == metric_name)
  if (nrow(plot_df) == 0) return(ggplot() + theme_void())

  models <- levels(plot_df$model_label)
  cols <- get_model_colours(models)
  if (is.null(y_label)) y_label <- metric_label(metric_name)

  ggplot(plot_df,
         aes(x = stratum, y = mean, fill = model_label, group = model_label)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.72,
             alpha = 0.85, colour = NA) +
    geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                  position = position_dodge(width = 0.8),
                  width = 0.28, linewidth = 0.35) +
    scale_fill_manual(values = cols, limits = models) +
    scale_y_continuous(limits = y_limits,
                       breaks = seq(y_limits[1], y_limits[2], 0.2),
                       expand = expansion(mult = c(0, 0.02))) +
    labs(title = title, x = NULL, y = y_label, fill = NULL) +
    theme_bw(base_size = 10) +
    theme(
      panel.grid.major.x = element_blank(),
      axis.text.x  = element_text(angle = 30, hjust = 1, size = 8),
      legend.position = "bottom",
      legend.text  = element_text(size = 8),
      plot.title   = element_text(face = "bold", size = 11, hjust = 0.5)
    ) +
    guides(fill = guide_legend(nrow = 1, byrow = TRUE))
}

STRATA_BAR_SPECS <- list(
  modality     = list(col = "modality",     levels = MODALITY_LEVELS,
                       pretty = FALSE, label = "modality"),
  content_type = list(col = "content_type", levels = CONTENT_TYPE_LEVELS,
                       pretty = FALSE, label = "content type"),
  location     = list(col = "location",     levels = ORGAN_LEVELS,
                       pretty = TRUE,  label = "organ")
)
```

```{r slake-ablation-bar-plots, fig.width=8.5, fig.height=5}
ablation_strata_bar_plots <- crossing(
  stratum = names(STRATA_BAR_SPECS),
  metric  = ablation_metrics,
  slice   = names(answer_type_slices)
) |>
  pmap(function(stratum, metric, slice) {
    spec <- STRATA_BAR_SPECS[[stratum]]
    slice_spec <- answer_type_slices[[slice]]
    sub_ps <- slice_spec$filter(abl_vqa_slake_ps)

    boot_df <- compute_strata_boot_cis(
      sub_ps, ablation_metrics, spec$col, spec$levels, pretty = spec$pretty
    )

    bar_with_ci_by_stratum(
      boot_df, metric,
      title = paste0("SLAKE ablation ", metric_label(metric),
                     " by ", spec$label, " - ", slice_spec$label)
    )
  })

walk(ablation_strata_bar_plots, print)
```
