---
title: "VReason reader study — finding retrieval vs rarity of a finding"
author: "Harry Robertson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6.5, fig.height = 4.2, dpi = 300)
```

```{r libs}
library(tidyverse)
library(lme4)
library(scales)

# ---- helpers (shared with reader-study-novel-findings.Rmd) -------------------
splitc <- function(x) {
  str_split(replace_na(as.character(x), ""), ";") |> unlist(use.names = FALSE) |>
    str_trim() |> (\(v) v[v != ""])()
}
# TRUE pathology burden (audit 2026-06-22, see reader-study-analysis/verification/). Exclude both
# "No Finding" (a token, not a pathology) and "Support Devices" (a device, not a retrievable
# pathology — also under-labeled in raw GT, and counting it leaked the normal case cb103846 into
# this analysis as a spurious very-rare "finding"). `pset()` is the pathology label SET per
# response; `npath()` its size.
NONPATH <- c("No Finding", "Support Devices")
pset  <- function(x) str_split(replace_na(as.character(x), ""), ";") |>
  map(~ setdiff(str_trim(.x[.x != ""]), NONPATH))
npath <- function(x) lengths(pset(x))
wilson <- function(k, n, z = 1.96) {                 # returns c(lo, hi) as proportion
  if (is.na(n) || n == 0) return(c(NA, NA))
  p <- k / n; d <- 1 + z^2/n
  c((p + z^2/(2*n))/d - z*sqrt(p*(1-p)/n + z^2/(4*n^2))/d,
    (p + z^2/(2*n))/d + z*sqrt(p*(1-p)/n + z^2/(4*n^2))/d)
}
P <- function(x) signif(x, 3)

COL_VR    <- "#8B1A1A"   # VReason red (reader + VReason)
COL_NEU   <- "#9CA3AF"   # neutral grey (unaided reader)
COL_AMBER <- "#E69F00"
COL_GREEN <- "#009E73"
COL_BLUE  <- "#0072B2"
theme_set(theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = rel(1.02)),
        plot.subtitle = element_text(colour = "grey35", size = rel(0.85)),
        legend.position = "bottom", legend.title = element_blank()))
```

A finding's **rarity** is its surprisal `-log(prevalence)` across the case set. The reader-study
question is whether *readers* miss rarer findings more, and whether **VReason recovers those rare
misses**. Working at the finding level in the **manual arm**: every ground-truth condition in every
reader×case response is one retrieval opportunity, scored as missed (unaided) and, if missed, as
recovered after VReason. The number of opportunities per condition is uneven (rare findings appear
in few cases), so inference comes from **GLMMs with `(1|user_id)+(1|case_id)`** and the descriptive
points carry Wilson intervals.

```{r data}
DL  <- "~/Downloads"
det <- read_csv(file.path(DL, "per_response_detailed.csv"), show_col_types = FALSE)

# Corrected case-level ground truth (audit 2026-06-22): drops 3 phantom "Enlarged
# Cardiomediastinum" labels (reports describe a normal mediastinum) and records omitted Support
# Devices. Keyed on full case_id; unmatched cases from a re-export fall back to raw via coalesce().
gtkey <- read_csv("reader-study-analysis/verification/gt_conditions_corrected.csv",
                  show_col_types = FALSE) |>
  select(case_id, gt_conditions_corrected)
det <- det |>
  left_join(gtkey, by = "case_id") |>
  mutate(gt_use = coalesce(gt_conditions_corrected, gt_conditions))

# prevalence / surprisal of each condition across ALL cases (pathology labels only)
Ncase <- n_distinct(det$case_id)
freq <- det |>
  transmute(case_id, cond = pset(gt_use)) |>
  unnest_longer(cond) |>
  distinct(case_id, cond) |>
  count(cond, name = "ncase") |>
  mutate(surprisal = -log(ncase / Ncase))

# manual-arm abnormal responses. Miss/recovery are recomputed from the corrected pathology sets
# (verified label-level: missed = gt-human, recovered = (gt-human) intersect humanai, 0/176
# mismatches), so the GT fix propagates; the burden guard is then satisfied by construction.
man <- det |>
  mutate(npath = npath(gt_use),
         missed_by_human = lengths(map2(pset(gt_use), pset(human_conditions), setdiff))) |>
  filter(presentation_mode_raw == "manual", npath > 0, missed_by_human <= npath)

# one row per (reader x case x ground-truth finding): was it missed? recovered by VReason?
cond_obs <- man |>
  transmute(case_id, user_id,
            cond = pset(gt_use),
            hum  = pset(human_conditions),
            hai  = pset(humanai_conditions)) |>
  unnest_longer(cond) |>
  mutate(was_missed   = !purrr::map2_lgl(cond, hum, ~ .x %in% .y),
         was_recov    = was_missed & purrr::map2_lgl(cond, hai, ~ .x %in% .y),
         still_missed = was_missed & !was_recov) |>
  left_join(freq, by = "cond")

# per-condition retrieval rates + Wilson intervals
ct <- cond_obs |> group_by(cond) |>
  summarise(n_obs = n(), n_missed = sum(was_missed), n_recovered = sum(was_recov & was_missed),
            surprisal = first(surprisal), ncase = first(ncase), .groups = "drop") |>
  mutate(h_caught   = n_obs - n_missed,
         hai_caught = n_obs - n_missed + n_recovered,
         human_recall   = h_caught / n_obs,
         humanai_recall = hai_caught / n_obs,
         recovery_rate  = ifelse(n_missed > 0, n_recovered / n_missed, NA_real_),
         h_lo   = purrr::map2_dbl(h_caught,   n_obs, ~ wilson(.x, .y)[1]),
         h_hi   = purrr::map2_dbl(h_caught,   n_obs, ~ wilson(.x, .y)[2]),
         hai_lo = purrr::map2_dbl(hai_caught, n_obs, ~ wilson(.x, .y)[1]),
         hai_hi = purrr::map2_dbl(hai_caught, n_obs, ~ wilson(.x, .y)[2])) |>
  arrange(surprisal)
```

```{r}
pneumonia_case_ids = cond_obs %>% filter(cond == "Pneumonia") %>% pull(case_id) %>% unique()

reviews <- read_csv(file.path("~/Downloads/finding_reviews_rows.csv"))
pneumonia_reviews = reviews %>% filter(case_id %in% pneumonia_case_ids)
```


---

## R1 — Readers miss rarer findings (headline)

Each point is a condition: unaided reader recall against rarity, sized by the number of retrieval
opportunities, with Wilson intervals. The GLMM tests the rarity→miss effect across the
finding-level observations, accounting for repeated readers and cases.

```{r R1, fig.height=4.6}
m_r1   <- glmer(was_missed ~ scale(surprisal) + (1|user_id) + (1|case_id),
                data = cond_obs, family = binomial, control = glmerControl(optimizer = "bobyqa"))
or_r1  <- exp(fixef(m_r1)["scale(surprisal)"]); p_r1v <- summary(m_r1)$coefficients["scale(surprisal)", 4]
rho1   <- cor(ct$surprisal, ct$human_recall, method = "spearman")

p_r1 <- ggplot(ct, aes(surprisal, human_recall)) +
  geom_smooth(method = "lm", colour = COL_NEU, fill = COL_NEU, alpha = 0.12) +
  geom_errorbar(aes(ymin = h_lo, ymax = h_hi), width = 0.04, colour = COL_NEU) +
  geom_point(aes(size = n_obs), colour = COL_NEU, alpha = 0.9) +
  geom_text(aes(label = cond), vjust = -0.9, size = 2.7, colour = "grey25") +
  scale_size_continuous(range = c(2, 6), name = "retrieval opportunities") +
  scale_y_continuous(labels = label_percent(), limits = c(-0.05, 1.15)) +
  labs(title = "R1: Unaided readers miss rarer findings",
       subtitle = sprintf("GLMM miss OR = %.2f per +1 SD rarity (p = %.3f); Spearman rho(recall) = %.2f",
                          or_r1, p_r1v, rho1),
       x = "Finding rarity  (surprisal = -log prevalence)", y = "Unaided reader recall")
p_r1
```

---

## R2 — The rarity gradient persists after VReason

The same recall against rarity, now overlaying unaided (grey) and reader+VReason (red). If VReason
preferentially rescued rare misses, the red slope would be flatter than the grey one — but the two
slopes are essentially parallel: VReason lifts recall roughly uniformly across rarity rather than
closing the gap at the rare end.

```{r R2, fig.height=4.6}
m_r2  <- glmer(still_missed ~ scale(surprisal) + (1|user_id) + (1|case_id),
               data = cond_obs, family = binomial, control = glmerControl(optimizer = "bobyqa"))
or_r2 <- exp(fixef(m_r2)["scale(surprisal)"])
rho2  <- cor(ct$surprisal, ct$humanai_recall, method = "spearman")

r2 <- ct |> select(cond, surprisal, n_obs, h_lo, h_hi, hai_lo, hai_hi,
                   Unaided = human_recall, `Reader + VReason` = humanai_recall) |>
  pivot_longer(c(Unaided, `Reader + VReason`), names_to = "arm", values_to = "recall") |>
  mutate(arm = factor(arm, levels = c("Unaided", "Reader + VReason")))

p_r2 <- ggplot(r2, aes(surprisal, recall, colour = arm, fill = arm)) +
  geom_smooth(method = "lm", alpha = 0.1) +
  geom_point(aes(size = n_obs), alpha = 0.85) +
  scale_colour_manual(values = c("Unaided" = COL_NEU, "Reader + VReason" = COL_VR), aesthetics = c("colour", "fill")) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  scale_y_continuous(labels = label_percent(), limits = c(-0.05, 1.1)) +
  labs(title = "R2: VReason does not flatten the rarity gradient",
       subtitle = sprintf("Rarity gradient essentially unchanged after VReason: still-missed OR = %.2f vs unaided %.2f per +1 SD rarity (rho %.2f vs %.2f)",
                          or_r2, or_r1, rho1, rho2),
       x = "Finding rarity  (surprisal = -log prevalence)", y = "Recall")
p_r2
```

---

## R3 — VReason recovery rate vs rarity

Of the findings readers missed unaided, the fraction VReason helped recover, by condition rarity.
Restricted to conditions that were missed at least once; point area is the number missed, bars are
Wilson intervals.

```{r R3}
r3 <- ct |> filter(n_missed > 0) |>
  mutate(rr_lo = purrr::map2_dbl(n_recovered, n_missed, ~ wilson(.x, .y)[1]),
         rr_hi = purrr::map2_dbl(n_recovered, n_missed, ~ wilson(.x, .y)[2]))
rho3 <- cor(r3$surprisal, r3$recovery_rate, method = "spearman")

p_r3 <- ggplot(r3, aes(surprisal, recovery_rate)) +
  geom_smooth(method = "lm", colour = COL_VR, fill = COL_VR, alpha = 0.12) +
  geom_errorbar(aes(ymin = rr_lo, ymax = rr_hi), width = 0.04, colour = COL_NEU) +
  geom_point(aes(size = n_missed), colour = COL_VR, alpha = 0.85) +
  geom_text(aes(label = cond), vjust = -0.9, size = 2.7, colour = "grey25") +
  scale_size_continuous(range = c(2, 6), name = "findings missed") +
  scale_y_continuous(labels = label_percent(), limits = c(-0.05, 1.18)) +
  labs(title = "R3: VReason recovery of missed findings vs rarity",
       subtitle = sprintf("Spearman rho = %.2f; rarest findings rest on 1-4 misses each (Wilson intervals)", rho3),
       x = "Finding rarity  (surprisal = -log prevalence)", y = "Share of unaided misses recovered")
p_r3
```

---

## R4 — Unaided → VReason-assisted recall, by condition

Per-condition dumbbells from unaided (grey) to reader+VReason (red), ordered rarest at top. The
segment length is the retrieval gain VReason delivers for each finding type.

```{r R4}
r4 <- ct |> mutate(cond = fct_reorder(cond, surprisal))

p_r4 <- ggplot(r4) +
  geom_segment(aes(y = cond, yend = cond, x = human_recall, xend = humanai_recall),
               colour = "grey70", linewidth = 1.1) +
  geom_point(aes(human_recall, cond, colour = "Unaided"), size = 3) +
  geom_point(aes(humanai_recall, cond, colour = "Reader + VReason"), size = 3) +
  geom_text(aes(humanai_recall, cond, label = sprintf("n=%d", n_obs)), hjust = -0.4, size = 2.6, colour = "grey35") +
  scale_colour_manual(values = c("Unaided" = COL_NEU, "Reader + VReason" = COL_VR)) +
  scale_x_continuous(labels = label_percent(), limits = c(0, 1.12)) +
  labs(x = "Finding recall", y = NULL)

ggsave("./figures/reader-study/rarity_R4.pdf", p_r4, width = 6.5, height = 4.2, device = "pdf")
```

---

> **Scope & caveats.** Manual arm only; finding-level observations from `r nrow(man)` responses on
> `r n_distinct(man$case_id)` abnormal cases. Opportunities per condition are very uneven (3–61) and
> the rarest findings rest on a handful of misses, so per-condition rates are uncertain (Wilson
> intervals) and the GLMMs — not the points — carry the inference. Rarity tracks finding count and
> difficulty, so this is the rarity axis of the reader study, not a rarity effect isolated from them.
> Recovery is observed only in the manual arm (`recovered_conditions` is empty by design in
> ai_assisted). Ground truth uses the corrected case-level key
> (`reader-study-analysis/verification/gt_conditions_corrected.csv`): "No Finding" and "Support
> Devices" are excluded from the finding set (the latter previously entered the rarity axis as a
> spurious ncase=1 condition), 3 phantom "Enlarged Cardiomediastinum" labels are dropped, and
> miss/recovery are recomputed from the corrected pathology sets.

```{r save, include=FALSE}
dir.create("./figures/reader-study", showWarnings = FALSE, recursive = TRUE)
walk2(list(p_r1, p_r2, p_r3, p_r4),
      paste0("./figures/reader-study/rarity_R", 1:4, ".pdf"),
      ~ ggsave(.y, plot = .x, width = 6.5, height = 4.2, device = "pdf", bg = "white"))
```
