---
title: "VReason reader study — finding retrieval vs number of findings"
author: "Harry Robertson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6.5, fig.height = 4.2, dpi = 300)
```

```{r libs}
library(tidyverse)
library(lme4)
library(scales)

# ---- helpers (shared with reader-study-novel-findings.Rmd) -------------------
splitc <- function(x) {
  str_split(replace_na(as.character(x), ""), ";") |> unlist(use.names = FALSE) |>
    str_trim() |> (\(v) v[v != ""])()
}
# TRUE pathology burden. Two verified GT labeling artifacts are excluded here (audit 2026-06-22,
# see reader-study-analysis/verification/gt_semantic_audit_2026-06-22.md):
#   (1) "No Finding" is a token, not a pathology;
#   (2) "Support Devices" is a device, not a pathology to be retrieved (it is also under-labeled in
#       the raw GT, present on only 1/12 device cases) — counting it leaked the normal case
#       cb103846 into the abnormal analysis.
# `pset()` returns the pathology label SET per response; `npath()` is its size. Working with the
# sets lets us recompute miss/recovery against the corrected GT in the data chunk below.
NONPATH <- c("No Finding", "Support Devices")
pset  <- function(x) str_split(replace_na(as.character(x), ""), ";") |>
  map(~ setdiff(str_trim(.x[.x != ""]), NONPATH))
npath <- function(x) lengths(pset(x))
wilson <- function(k, n, z = 1.96) {                 # returns c(lo, hi) as proportion
  if (is.na(n) || n == 0) return(c(NA, NA))
  p <- k / n; d <- 1 + z^2/n
  c((p + z^2/(2*n))/d - z*sqrt(p*(1-p)/n + z^2/(4*n^2))/d,
    (p + z^2/(2*n))/d + z*sqrt(p*(1-p)/n + z^2/(4*n^2))/d)
}
P <- function(x) signif(x, 3)

COL_VR    <- "#8B1A1A"   # VReason red (reader + VReason)
COL_NEU   <- "#9CA3AF"   # neutral grey (unaided reader)
COL_AMBER <- "#E69F00"   # Okabe-Ito orange
COL_GREEN <- "#009E73"   # Okabe-Ito green
COL_BLUE  <- "#0072B2"   # Okabe-Ito blue
theme_set(theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = rel(1.02)),
        plot.subtitle = element_text(colour = "grey35", size = rel(0.85)),
        legend.position = "bottom", legend.title = element_blank()))
```

This is a **reader-study** analysis: the outcome is how many ground-truth findings the *readers*
retrieve, unaided and then with VReason. We use the **manual arm** (read unaided first, VReason
revealed second), where `missed_by_human` is the unaided miss count and `recovered_by_AI` is the
subset of those misses the reader caught after seeing VReason. Each row is one reader×case response;
cases carry an uneven number of readers (1–7), so all inferential statements come from binomial
**GLMMs with `(1|user_id)+(1|case_id)`** rather than from collapsing to the case.

```{r data}
DL  <- "~/Downloads"
det <- read_csv(file.path(DL, "per_response_detailed.csv"), show_col_types = FALSE)

# Corrected case-level ground truth (audit 2026-06-22): removes 3 phantom "Enlarged
# Cardiomediastinum" labels whose reports describe a normal mediastinum, and records the
# Support Devices the raw GT omitted. Keyed on full case_id; new cases from a re-export that are
# not yet in the key fall back to the raw gt_conditions via coalesce().
gtkey <- read_csv("reader-study-analysis/verification/gt_conditions_corrected.csv",
                  show_col_types = FALSE) |>
  select(case_id, gt_conditions_corrected)

# The miss/recovery counts in the export are exactly  gt − human  and  (gt − human) ∩ humanai
# (verified label-level, 0/176 mismatches), so we recompute them from the CORRECTED pathology
# sets rather than trusting the precomputed columns — this propagates the GT fix into the counts.
man <- det |>
  left_join(gtkey, by = "case_id") |>
  mutate(gt_use = coalesce(gt_conditions_corrected, gt_conditions),
         gt_p   = pset(gt_use),
         hum_p  = pset(human_conditions),
         hai_p  = pset(humanai_conditions),
         npath           = lengths(gt_p),
         missed_by_human = lengths(map2(gt_p, hum_p, setdiff)),
         recovered_by_AI = lengths(map2(map2(gt_p, hum_p, setdiff), hai_p, intersect))) |>
  filter(presentation_mode_raw == "manual", npath > 0, missed_by_human <= npath) |>
  mutate(caught_h   = npath - missed_by_human,                       # unaided catches
         caught_hai = pmin(npath, caught_h + recovered_by_AI),       # catches after VReason
         h_recall   = caught_h / npath,
         hai_recall = caught_hai / npath,
         miss_rate  = missed_by_human / npath,
         bin        = factor(pmin(npath, 3), levels = 1:3, labels = c("1", "2", "3+")))

```

---

## C1 — Unaided vs VReason-assisted recall, by finding count

Share of ground-truth findings the reader retrieves, in finding-count bands, unaided (grey) vs after
VReason (red). Bars are finding-level Wilson 95% intervals (descriptive); the burden trend is tested
with a GLMM in the next panels.

```{r C1}
c1 <- man |> group_by(bin) |>
  summarise(gt = sum(npath), H = sum(caught_h), HAI = sum(caught_hai), n_resp = n(), .groups = "drop") |>
  pivot_longer(c(H, HAI), names_to = "arm", values_to = "caught") |>
  mutate(recall = caught / gt,
         lo = purrr::map2_dbl(caught, gt, ~ wilson(.x, .y)[1]),
         hi = purrr::map2_dbl(caught, gt, ~ wilson(.x, .y)[2]),
         arm = factor(arm, levels = c("H", "HAI"),
                      labels = c("Unaided reader", "Reader + VReason")))
ov_h   <- 100 * sum(man$caught_h)   / sum(man$npath)
ov_hai <- 100 * sum(man$caught_hai) / sum(man$npath)

p_c1 <- ggplot(c1, aes(bin, recall, fill = arm)) +
  geom_col(position = position_dodge(0.7), width = 0.62, alpha = 0.9) +
  scale_fill_manual(values = c("Unaided reader" = COL_NEU, "Reader + VReason" = COL_VR)) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.5)) +
  labs(x = "Ground-truth findings per case  (npath)", y = "Findings retrieved")
p_c1

ggsave("./figures/reader-study/findingcount_C1.pdf", plot = p_c1, width = 6.5, height = 4.2, device = "pdf", bg = "white")
```

---

## C2 — Unaided miss rate rises with finding burden

Per response, the fraction of true findings the unaided reader missed, against the finding count.
The curve is a binomial fit; the odds ratio is from a GLMM that accounts for repeated readers and
cases.

```{r C2}
m_c2 <- glmer(cbind(missed_by_human, caught_h) ~ scale(npath) + (1|user_id) + (1|case_id),
              data = man, family = binomial, control = glmerControl(optimizer = "bobyqa"))
sc   <- summary(m_c2)$coefficients
or_c2 <- exp(sc["scale(npath)", 1]); p_c2v <- sc["scale(npath)", 4]

p_c2 <- ggplot(man, aes(npath, miss_rate)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              aes(weight = npath), colour = COL_VR, fill = COL_VR, alpha = 0.12) +
  geom_jitter(width = 0.12, height = 0.02, alpha = 0.5, colour = COL_NEU, size = 1.5) +
  scale_x_continuous(breaks = 1:5) +
  scale_y_continuous(labels = label_percent(), limits = c(-0.05, 1.05)) +
  labs(title = "C2: Unaided miss rate is flat across finding burden",
       subtitle = sprintf("GLMM miss OR = %.2f per +1 SD of finding count (p = %.3f, n.s.); readers miss ~3 in 4 findings regardless", or_c2, p_c2v),
       x = "Ground-truth findings per case  (npath)", y = "Unaided miss rate")
p_c2
```

---

## C3 — VReason recovery of missed findings, by burden

Of the findings the reader missed unaided, what fraction did VReason help recover? Restricted to
responses with at least one miss; point area is the number missed.

```{r C3}
c3 <- man |> filter(missed_by_human > 0) |>
  mutate(rec_rate = recovered_by_AI / missed_by_human)
m_c3 <- glmer(cbind(recovered_by_AI, missed_by_human - recovered_by_AI) ~ scale(npath) +
                (1|user_id) + (1|case_id),
              data = c3, family = binomial, control = glmerControl(optimizer = "bobyqa"))
sc3 <- summary(m_c3)$coefficients
or_c3 <- exp(sc3["scale(npath)", 1]); p_c3v <- sc3["scale(npath)", 4]

p_c3 <- ggplot(c3, aes(npath, rec_rate)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              aes(weight = missed_by_human), colour = COL_VR, fill = COL_VR, alpha = 0.12) +
  geom_jitter(aes(size = missed_by_human), width = 0.12, height = 0.02, alpha = 0.5, colour = COL_NEU) +
  scale_size_continuous(range = c(1.5, 5), name = "findings missed") +
  scale_x_continuous(breaks = 1:5) +
  scale_y_continuous(labels = label_percent(), limits = c(-0.05, 1.05)) +
  labs(title = "C3: VReason recovery rate is roughly constant across burden",
       subtitle = sprintf("GLMM recovery OR = %.2f per +1 SD of finding count (p = %.3f, n.s.)", or_c3, p_c3v),
       x = "Ground-truth findings per case  (npath)", y = "Share of missed findings recovered")
p_c3
```

---

## C4 — The assistance benefit per response shrinks with burden

The within-response recall gain from VReason (`reader+VReason − unaided`), by finding-count band. The
dashed line is no benefit; the boxes show the gain is largest on low-burden cases.

```{r C4}
c4 <- man |> mutate(gain = hai_recall - h_recall)
gain_by_bin <- c4 |> group_by(bin) |> summarise(mg = mean(gain), n = n(), .groups = "drop")

p_c4 <- ggplot(c4, aes(bin, gain)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_boxplot(width = 0.55, alpha = 0.4, fill = COL_VR, outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0.01, alpha = 0.5, size = 1.4) +
  geom_text(data = gain_by_bin, aes(bin, 1.02, label = sprintf("+%.0f%%\n(n=%d)", 100*mg, n)),
            size = 3, lineheight = 0.9, colour = "grey25") +
  scale_y_continuous(labels = label_percent(), limits = c(-0.2, 1.1)) +
  labs(title = "C4: Per-case recall gain from VReason, by finding burden",
       subtitle = "Gain = (reader + VReason) − unaided recall, within the same response",
       x = "Ground-truth findings per case  (npath)", y = "Recall gain from VReason")
p_c4
```

---

> **Scope & caveats.** Manual arm only (the unaided→assisted crossover); `r nrow(man)` responses on
> `r n_distinct(man$case_id)` abnormal cases. Readers per case are uneven (1–7), which is exactly why
> the inferential claims use `(1|user_id)+(1|case_id)` GLMMs rather than case-collapsed rates — the
> bar/point CIs are descriptive (finding-level Wilson). Finding count is near-collinear with the
> difficulty label (N5 in `reader-study-novel-findings.Rmd`), so read these as the burden axis, not a
> burden effect isolated from difficulty. Ground truth uses the corrected case-level key
> (`reader-study-analysis/verification/gt_conditions_corrected.csv`): "No Finding" and "Support
> Devices" are excluded from the pathology burden, and 3 phantom "Enlarged Cardiomediastinum"
> labels are dropped; miss/recovery counts are recomputed from the corrected pathology sets and
> guarded to the burden (`missed_by_human ≤ npath`).

```{r save, include=FALSE}
dir.create("./figures/reader-study", showWarnings = FALSE, recursive = TRUE)
walk2(list(p_c1, p_c2, p_c3, p_c4),
      paste0("./figures/reader-study/findingcount_C", 1:4, ".pdf"),
      ~ ggsave(.y, plot = .x, width = 6.5, height = 4.2, device = "pdf", bg = "white"))
```
